---
title: "Stat 897 Fall 2017 Data Analysis Assignment 8"
author: "Penn State"
date: "Due October 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this assignment we again use the College data found in the ISLR library, with the 600 observation training set (using the rest as the test data).

### (a) Using out-of-state tuition as the response and the other variables as the predictors, perform forward stepwise selection on the training set in order to identify a satisfactory model that uses just a subset of the predictors. Please set the seed at 801 once at the beginning before choosing the training set.

```{r}
library(leaps)
library(ISLR)
#install.packages('ISLR')
library(gam)
data("College")
attach(College)

set.seed (801)
trainingRows=sample (nrow(College), 600, replace = FALSE)
train = College[trainingRows,]
test = College[-trainingRows,]

regfit.fwd=regsubsets (Outstate~.,data=train, nvmax =17, method='forward')
reg.summary = summary (regfit.fwd)
plot(reg.summary$bic, xlab ="Number of Variables",ylab="BIC", type = 'l', main = 'Forward Step - Performance Measure')
which.min (reg.summary$bic )
points (which.min (reg.summary$bic ), reg.summary$bic[which.min (reg.summary$bic )], col ="red",cex =2, pch =20)

plot(reg.summary$cp, xlab ="Number of Variables",ylab="cp", type = 'l', main = 'Forward Step - Performance Measure')
which.min (reg.summary$cp )
points (which.min (reg.summary$cp ), reg.summary$cp[which.min (reg.summary$cp )], col ="red",cex =2, pch =20)

plot(reg.summary$adjr2, xlab ="Number of Variables",ylab="Adj R2", type = 'l', main = 'Forward Step - Performance Measure')
which.max (reg.summary$adjr2 )
points (which.max (reg.summary$adjr2 ), reg.summary$cp[which.max (reg.summary$adjr2 )], col ="red",cex =2, pch =20)

test.mat=model.matrix (Outstate~.,data=test)
test.val.errors =rep(NA ,17)
for(i in 1:17){
	coefi=coef(regfit.fwd ,id=i)
	pred=test.mat [,names(coefi)] %*% coefi
	test.val.errors [i]= mean(( test$Apps-pred)^2)
}
plot(test.val.errors ,type='b', xlab='# of parameters', ylab='Test MSE')

```

The above indicates that we get the following choice based on the different measures:

BIC - 6
CP - 13
Adjusted R2 - 14

From a visual inspection, it is clear that the elbow in all the plots happens at 6 and therefore it could be a good choice to move forward. Let see the various parameters included in this model

```{r}
names(coef(regfit.fwd ,id=6))
```

### (b)  Fit a GAM on the training data, using out-of-state tuition as the response and the features selected in the previous step as the predictors. Use appropriate nonlinear components (e.g. natural splines, step functions) for the variables that need it as in the salary example from the book.  Plot the results, and explain your findings.   What nonlinear components did you use?

```{r}
#"PrivateYes"  "Room.Board"  "Terminal"    "perc.alumni" "Expend"      "Grad.Rate"
fit.rb= lm(Outstate~poly(Room.Board ,5) ,data=train)
coef(summary(fit.rb))
# The relationship appears to be only linear

fit.t= lm(Outstate~poly(Terminal ,5) ,data=train)
coef(summary(fit.t))
# We have a cubic relationship, we can try ns

fit.pa= lm(Outstate~poly(perc.alumni ,5) ,data=train)
coef(summary(fit.pa))
# The relationship appears to be only linear

fit.e= lm(Outstate~poly(Expend ,5) ,data=train)
coef(summary(fit.e))
# We have a cubic relationship, we can try ns

fit.gr= lm(Outstate~poly(Grad.Rate ,5) ,data=train)
coef(summary(fit.gr))
# We have a cubic relationship, we can try ns

fit.1=gam(Outstate~Room.Board, data=train)
fit.2=gam(Outstate~Room.Board+poly(Terminal,3), data=train)
anova(fit.1, fit.2)
# 2nd is better, lets continue

fit.3=gam(Outstate~Room.Board+poly(Terminal,3)+perc.alumni, data=train)
anova(fit.1, fit.2, fit.3)
# 3rd is best, lets continue

fit.4=gam(Outstate~Room.Board+poly(Terminal,3)+perc.alumni+poly(Expend,3), data=train)
anova(fit.1, fit.2, fit.3, fit.4)
# 4th is best, lets continue

fit.5=gam(Outstate~Room.Board+poly(Terminal,3)+perc.alumni+poly(Expend,3)+poly(Grad.Rate,3), data=train)
anova(fit.1, fit.2, fit.3, fit.4, fit.5)
# 5th is best, lets continue

fit.6=gam(Outstate~Room.Board+poly(Terminal,3)+perc.alumni+poly(Expend,3)+poly(Grad.Rate,3)+Private, data=train)
anova(fit.1, fit.2, fit.3, fit.4, fit.5, fit.6)

# We will try a model with natural splines
fit.7=gam(Outstate~Room.Board+ns(Terminal)+perc.alumni+ns(Expend)+ns(Grad.Rate)+Private, data=train)
anova(fit.6, fit.7)
# Model with ns performs better

par(mfrow =c(3,3))
plot(fit.7, se=TRUE ,col ="blue ")

pred=predict (fit.7 ,newdata =test, se=T)
mean((test$Outstate - pred$fit)^2)

```

### (c) Evaluate both models obtained from part (a) and (b) on the test set, and explain the results obtained.


```{r}
pred=predict (fit.7 ,newdata =test, se=T)
mean((test$Outstate - pred$fit)^2)

```

### (d) For which variables, if any, is there evidence of a non-linear relationship with the response?

We reviewed all these above individually and found non-linear relationships for the following: Terminal, Expend, Grad.Rate

```{r}

```